State of the Practice for Lattice Boltzmann Method Software

### 2022-03-01 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to section 10 on recommendations

Requirements documentation is feasible, as shown by the prevalence of theory
documents in LBM software.  Table~\ref{artifactspresent} shows that theory
documentation is common.  We observed 17 of 24 packages had at least some theory
documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
SmithAndLai2005}, show theory is a significant part of the requirements
documentation. With the addition of some extra information, the theory documents
can be transformed into requirements specification.  The key extra information
includes explicitly stating user characteristics, explicitly stating how the
user interacts with the software in terms of input data requirements, and
listing likely and unlikely changes.  Explicit statements about likely changes
are invaluable in the design stage, since they provide developers guidance on
how general the software needs to be.  For those wishing to maximize the value
of a requirement specification, information could be added like prioritizing the
nonfunctional requirements (to show the relative importance between qualities
like portability, reliability and performance) and traceability information (to
show the consequences of any changes to the assumptions).

@@

\subsection{Prioritize Documentation} \label{Sec_RecommendDoc}

Despite it being a pain point, spending time on documentation provides a
justifiable return on investment.  Addresses \ref{P_Documentation} on
documentation.

Software packages were not quantitatively measured for traceability. In this
project we asked developers to comment on traceability, specifically on their
software package's documentation and how it fit into their development process.

The interviewed developer of ESPResSo noted that all major additions to their
package had accompanying changes to artifacts and documentation. They noted that
considerable effort had been put into the documentation. They further commented
that they want to lower the entry barrier for new developers, and because of
that their package has a considerable amount of developer documentation. This
documentation informs developers on how to get started, and orients them to the
artifacts, source code, and system architecture, as well as how the software
package build system works, and how the coupling between the simulation engine
and the interface works. \wss{If ESPresSo scored high (I think it did), we can
use that to further promote the value of documentation.  Also a threat to
validity is that we have implicitly assumed documentation is valuable when
measuring quality.}

Finding ways to hasten updating documentation would increase the frequency of
such updates and benefit both users and developers.  \wss{Same thing has come up
in papers.  I'll try to find a reference for this.  This is also something we
can point to for recommendations related to code generation.} The developer of
OpenLB noted the use of documentation generators like Doxygen. It would be
advisable for more projects to use such automatic document generation tools,
since some projects do not do this. \wss{We can refer back to this for the
recommendations.}

Reusability. Provide API documentation, if applicable. Only one (ESPResSo) of
the top five ranked packages provided API documentation.

\wss{Address Usability (\ppref{P_Usability}): Better documentation, with
attention on traceability, and automatic code generation are suggested when
designing software for change, and would help with these modifications.  So far
this problem of on-boarding new users has been addressed by updating the
documentation to better inform users of the underlying LBM theory and package
requirements. Similar issues with LBM parameters were noted by another
developer. Updating the user interface to better explain theoretical principles,
as well as test user input for compatibility, was the implemented solution. As
noted above, sometimes frequently asked questions on the underlying theory and
on how to use the software are answered in the documentation.  The interviewed
developer of ESPResSo commented that parts of their package's source code had
been refactored to Python to help address usability issues. Python was perceived
as a much more usable language, and it would be easy for future users and
developers to learn and understand the source code.}

One software package (ESPResSo) changed some of its scripting language to Python
to make it more usable. The developer commented that this was ``the biggest step
in terms of usability over the years'', further commenting that ``most people in
the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
approach facilitates switching to whatever language is currently popular.}

\begin{itemize}
\item Usability. Understandability. Unambiguity. State appropriate fluid
dynamics problems that the software is designed to model in the documentation,
and explicitly state the limits of the software. This is done in the ESPResSo
user guide.
\item Usability. Provide documentation that details the background theory
information, or provide a reference to such information. This was done by
several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
\item Usability. Identify expected user characteristics. LIMBES, Ludwig,
laboetie, and Palabos did this. The importance of specifying user
characteristics is discussed in \citep{smith2007requirements}.
\item Usability. Keep all documentation in one location. This was done by top
ranked packages.
\item Usability. Do not duplicate artifact information. This is suggested as
good software development practice. Duplicate information is difficult to
maintain. 
\end{itemize}

@@

Borrow from SE for HEP paper.

Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
approach is to list the practice and then the PPs that it helps.  Our
recommendations are not lists of what should have been done in the past, or what
should be done now; they are just suggestions for consideration in the future.

``unique challenges of scientific research (e.g., the frequent and unforeseen
changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
the future this viewpoint could change and why.  Not really a unique challenge.
Cite Smith 2016?

linters \citep{Zadka2018} were rarely observed in LBM projects.  Do not see much evidence of their presence in best practice guides.

\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
on-boarding new developers, time constraints could be mitigated by increasing
developer efficiency, which could be addressed in several ways, including by
improving the quality of documentation, or incorporating automatic code
generation.}

\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
build confidence in correctness.  Peer review.  }

\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
commented that there are no journals that publish such research software
source code. However, there are ways to get such source code cited. Work has
been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
set of software citation principles and discusses ``how they could be used to
implement software citation in the scholarly community''
\citep{katz2019software}. other papers also mention this pain point.
Solutions proposed by Katz, journal of open source etc. Journal of Open Source
Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}

\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
background of users explicit, possibly in a requirements document.}

@@

Test the installation and tutorials.  Assign a junior team member the task?

Papers that talk about how to do open source.

\wss{in recommendations could point out what teams could do to get more external
contributors, and potentially a longer life for their project.}

- previous section discusses challenges with requirements and not popular (maybe
move that section here) - say what information to incorporate, low hanging fruit
- return on investment

\subsection{Verifiability}

- further steps to address the oracle problem - borrow from SE for HEP paper.
- anything in Smith 2016?

\subsection{Reproducibility}

- record versions of everything
- Docker, Virtual machines

@@

@@

@@

\subsection{Future State Of The Practice Assessments} \label{futuresop}

@@

Recommendations for future state of the practice assessments are made in Section
~\ref{futuresop}.

@@

- data on Mendeley

- make recommendations in order of conservative to ``bleeding edge''
- add pointer to project checklists (as sent to Jacques, as promoted during workshop)

@@

>>>>

@@

\subsection{Prioritize Documentation and Usability} \label{Sec_RecommendDoc}

Although documentation is difficult (as mentioned under pain point
\ref{P_Documentation}), some LBM developers have realized that prioritizing time
on documentation provides a justifiable return on investment. For instance, the
interviewed developer of ESPResSo noted that all major additions to their
package had accompanying changes to artifacts and documentation. They noted that
considerable effort was put into their documentation. They further commented
that their goal was to lower the entry barrier for new developers, which meant
providing considerable developer documentation. This documentation informs
developers on how to get started; orients them to the artifacts, source code,
system architecture, and build system; and, explains the coupling between
the simulation engine and the interface. In addition, ESPResSo developers
included API documentation, although they are the only one of the top five
packages to do so.

Furthermore, documentation is used by LBM developers to address usability
challenges (\ref{P_Usability}), as follows:

\begin{itemize}
	\item Explicitly state appropriate fluid dynamics problems that the software
	is designed to model and explicitly state the limits of the software. This
	is done in the ESPResSo user guide.
	\item Provide documentation that details the background theory, or provide a
	reference to such information. This was done by several top ranked packages,
	including ESPResSo, Ludwig, LUMA, and OpenLB.
	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and
	Palabos all did this. Explicit user characteristics are important so the
	application and documentation can be pitched toward the right user
	background \citep{smith2007requirements}.
	\item Keep all documentation in one location. This was done by all top
	ranked packages.
	\item Use documentation generators, like Doxygen, as done (for instance) by
	OpenLB.  More details on the packages that use generators is given in
	Sections~\ref{SecSurfCorrectAndVerifiab} and~\ref{Sec_CompareTools}.
	\item Provide an explanation of theoretical principles through the user interface.
	\item Add guards in the code to test user inputs for compatibility before
	proceeding with the calculations.
	\item Include documentation on frequently asked questions (Section~\ref{Sec_CompareArtifacts}).
\end{itemize}

@@

In this section we provide recommendations to address the pain points from
Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
are not lists of criticisms what should have been done in the past, or what
should be done now; they are just suggestions for consideration in the future.
We will not be repeating ideas here that have been discussed in previous
section, such as continuous integration, documentation of APIs, etc.  Our aim is
to mention ideas that are at least somewhat beyond conventional best practices.
The ideas listed have the potential to become best practices in the medium to
long-term.  The ideas are provided roughly in the order of increasing novelty.

\subsection{Employ Linters} \label{Sec_Linters}

With the exception of \citet{ThielEtAl2020}, the research software guidelines
that we consulted do not mention linters.  However, we feel linters have the
potential to improve code quality at a relatively low cost.  A linter is a tool
that statically analyzes code to find programming errors, stylistic
inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
can be used to spot check code files, or even better as part of a continuous
integration system.

A sample of the potential benefits of linters include: finding memory leaks,
finding potential bugs, standardizing code with respect to formatting, improving
performance, removing silly errors before code reviews, and catching potential
security issues \citep{SourceLevel2022_Lint}.  Linters are available for most
popular programming languages.  For instance Python has the options of PyLint,
flake8 and Black \citep{Zadka2018}.

Linters can address the pain points for LBM developers.  For instance, a linter
can increase the amount of development time (\ppref{P_LackDevTime}) by
decreasing the number of mundane mistakes programmers have to catch.  Since the
linter can include rules that capture the wisdom of senior programmers, it can
help newer developers avoid common mistakes (\ref{P_LackSoftDevExp}).  Although
there do not appear to be many linters for checking parallel programming
mistakes (\ppref{P_TechnologyHurdles}), Parallel Lint is an option for OpenMP.
If a linter is used consistently this can guard against some technical debt
(\ppref{P_TechDebt}). Although linters are thought of as tools for code
analysis, similar ideas can be applied to documentation to ensure that at least
the basic documentation rules are followed.  The use of tools to check
documentation is a partial explanation for the relatively higher quality of
statistical tools that are part of the Comprehensive R Archive Network (CRAN)
\citep{SmithEtAl2018_StatSoft}.

\subsection{Conduct Rigorous Peer Reviews}

Peer review is already a strategy adopted by some LBM developers to improve
confidence in their software (Section~\ref{SecReproducibility}), but the
benefits of peer review can be improved by making its application more rigorous.
\citet{Jones2008} shows that rigorous inspection finds 60-65\% of latent defects
on average, and often tops 85\% in defect removal efficiency.  By way of
comparison, most forms of testing average between 30 and 35\% for defect removal
efficiency \citep{EbertAndJones2009, Jones2008}.  A formal code inspection
involves asking the reviewer to either follow a review checklist (check
consistency of variable names, look for terminating loops, etc.), or perform
specific review tasks (like summarize the purpose of the code, create a data
dictionary for the module assigned, cross reference the code to the technical
manual, etc.) The task based inspection approach has been effectively used for
research software, as described by \citet{KellyAndShepard2000}. Task based
inspection is an ideal fit with an issue tracking system, like GitHub.  The
review tasks can be issues, so that they can be easily assigned, monitored and
recorded.  Other potential issues for the tracker include assigning junior team
members to test installation instructions and getting-started tutorials.

Rigorous peer review addresses the same pain points as linters
(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ref{P_LackSoftDevExp},
\ppref{P_TechnologyHurdles}), \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
review addresses these pain points by efficiently searching for defects and
problems.

\subsection{Write and Submit More Papers on Software}

To address the pain point of a lack of funding (\ppref{P_LackFunding}), LBM
developers should watch for new opportunities to publish their research software
source code. For instance, the Journal of Open Source Software (JOSS) reviews
and publishes articles that credit the scholarship contained in the software
itself \citep{SmithEtAl2018-joss}.  \citep{SmithEtAl2016-softcite} presents a
set of software citation principles that may encourage broad adoption of a
consistent policy for software citation across disciplines and venues.
\citep{ChueHongEtAl2019} provides a software citation checklist for developers
to make a release of their software citable.  Further guidance on software
citation is provided by \citet{KatzEtAl2021}.  Another option is to use GitHub
for citation, since it is relatively easy to add metadata to repositories and to
generate citations for those repositories \citet{Smith2021-CitationsOnGitHub}.

Addresses lack of development time (\ppref{P_LackDevTime}) by increasing the
number of developers.

Papers that talk about how to do open source. \citep{Jalan2016}

Get more developers by better documentation, tutorials, etc.

Combine existing teams?

\subsection{Further Address the Oracle Problem} \label{Sec_FurtherOracle}

- say what PPs are addressed
- further steps to address the oracle problem - borrow from SE for HEP paper.
- anything in Smith 2016?
- introduce list on page 11 of HEP paper, add repeatability point

In the absence of a test oracle, the following
techniques can be used to build system tests \citep{Smith2016}:

\begin{itemize}

\item Test cases can be selected that are a subset of the real problem for which
  a closed-form solution exists.  When using this approach, confidence in the
  actual production code can only be built if it is the same code used for
  testing; that is, nothing is gained if a separate, simpler, program is written
  for testing the special cases.
\item Verification test cases can be created by assuming a solution and using
  this to calculate the inputs.  For instance, for a linear solver, if $A$ and
  $x$ are assumed, $b$ can be calculated as $b = A x$.  Following this, $A x^* =
  b$ can be solved and then $x$ and $x^*$, which should theoretically be equal,
  can be compared.  In the case of solving Partial Differential Equations
  (PDEs), this approach is called the Method of Manufactured Solutions
  \citep{Roache1998}.
\item Most scientific software uses floating point arithmetic, but for testing
  purposes, the slower, but guaranteed correct, interval arithmetic
  \citep{Hickey2001} can be employed. The faster floating point algorithm can
  then be verified by ensuring that the calculated answers lie within the
  guaranteed bounds.
\item Verification tests should include plans for convergence studies.  The
  discretization used in the numerical algorithm should be decreased (usually
  halved) and the change in the solution assessed.
\item Confidence can be built in a numerical algorithm by comparing the results
  to another program that overlaps in functionality.  If the test results do not
  agree, then one, possibly both of the programs is incorrect.
\item The verification plan should also include test plans for nonfunctional
  requirements, like accuracy, performance and portability, if these are
  important implementation goals.  Performance tests can be planned to describe
  how the software responds to changing inputs, such as problem size, condition
  number etc.  Verification plans can include relative comparisons between
  the new implementation and competing products \citep{Smith2006}.
\item In cases where a test oracle is not available, as is usually the case for
  scientific software, \emph{metamorphic} testing can be used.  ``This technique
  operates by checking whether the program behaves according to a certain set of
  properties called metamorphic relations.  A metamorphic relation is a
  relationship between multiple inputs and output pairs of the program.''
  \citep{KanewalaAndLundgren2016}.
\end{itemize}

Point to Section~\ref{Sec_CompareArtifacts}. Requirements documentation is
feasible, as shown by the prevalence of theory documents in LBM software.
Table~\ref{artifactspresent} shows that theory documentation is common.  We
observed 17 of 24 packages had at least some theory documentation. Templates for
scientific requirements, like \citet{SmithEtAl2007, SmithAndLai2005}, show
theory is a significant part of the requirements documentation. With the
addition of some extra information, the theory documents can be transformed into
requirements specification.  The key extra information includes explicitly
stating user characteristics, explicitly stating how the user interacts with the
software in terms of input data requirements, and listing likely and unlikely
changes.  Explicit statements about likely changes are invaluable in the design
stage, since they provide developers guidance on how general the software needs
to be.  For those wishing to maximize the value of a requirement specification,
information could be added like prioritizing the nonfunctional requirements (to
show the relative importance between qualities like portability, reliability and
performance) and traceability information (to show the consequences of any
changes to the assumptions).

- say what PPs are addressed
- previous section discusses challenges with requirements and not popular (maybe
move that section here) - say what information to incorporate, low hanging fruit
- return on investment
- maybe borrow from SE for HEP
- toward goal of replicability \citep{BenureauAndRougier2017}.

Addresses lack of development time (\ppref{P_LackDevTime}) by getting developers
up to speed more quickly.  Reducing the amount of time needed to train new
developers.

@@

\subsection{Improve Repeatability, Reproducibility and Replicability}

- say what PPs are addressed
- record versions of everything
- Docker, Virtual machines

@@

\subsection{Generate All Things}

- say what PPs are addressed Addresses lack of development time
(\ppref{P_LackDevTime}) by automating the writing of documentation and code.

@@

a threat to validity is that we have implicitly assumed documentation is
valuable when measuring quality.

threat to validity is the person doing the measuring.  In the previous exercise
we did do an experiment to measure this and convince ourselves that it isn't
that significant a risk.

@@

Recommendations, starting from current best practices.

\begin{enumerate}
	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.  Review the practices from the previous section.  Addresses PP?.
	\item \textbf{Continuous integration.} Addresses PP?.
	\item \textbf{Document API.} Addresses PP?.
\end{enumerate}

Future State Of The Practice Assessments

@@

\section*{Acknowledgements}

We would like to thank Ao Dong and Oluwaseun Owojaiye for many fruitful
discussions on topics relevant to this paper.

@@

@@

- ``unique challenges of scientific research (e.g., the frequent and unforeseen
changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
the future this viewpoint could change and why.  Not really a unique challenge.
Cite Smith 2016?
